\thetitle

### 2020-10-23 sun_phases/spin_qudit_tomo.tex
minor wording changes

For the practically minded, the mere existence of solutions to a problem is less interesting than the exposition of a particular solution.
On a high level, a spin qudit tomography protocol needs to
\begin{enumerate*}
\item select a set of measurement axes,
\item collect measurement data on spin projection onto these axes, and then
  \label{tomo:measure}
\item process the collected data to reconstruct the state of the spin qudit.
\end{enumerate*}
Whereas step \ref{tomo:measure} can involve a host of platform-dependent technical challenges, here we discuss the steps to take before and after measurement data collection.

@@

A set of measurement axes $V=\set{\v v}$ nominally induces a set of projectors $\v\Pi_V=\set{\Pi_j^V}$ that will be measured in an experiment, where for shorthand we use a combined index $j\simeq\p{\v v_j,\mu_j}$ that specifies both a measurement axis $\v v_j$ and an outcome $\mu_j$, and for clarify we index objects by $V$ to emphasize their dependence on the choice of measurement axes.
The projectors $\v\Pi_V$ can be used to construct the overlap (Gram) matrix
\begin{align}
  G_V = \sum_{i,j=1}^{\abs{V}\times d} \obk{\Pi_i^V|\Pi_j^V} \op{i}{j}.
  \label{eq:overlap_matrix}
\end{align}
Eigenvectors of $G_V$ define linearly independent linear combinations of the projectors in $\v\Pi_V$.
If the projectors in $\v\Pi_V$ span the entire space of operators on a $d$-level spin qudit, which is a necessary and sufficient condition for $V$ to allow for full state tomography, then $G_V$ must have exactly $d^2$ nonzero eigenvalues.
Indexing these eigenvalues $G^V_k$ and the corresponding (normalized) eigenvectors $\v x^V_k\equiv(x^V_{k,1},x^V_{k,2},\cdots)$ by an integer $k\in\set{1,2,\cdots,d^2}$,
we can construct the orthonormal operators
\begin{align}
  Q^V_k \equiv \sum_j g^V_{kj} \Pi_j^V,
  &&
  g^V_{kj} \equiv \f{x^V_{kj}}{\sqrt{G^V_k}}.
\end{align}
These operators allow us to expand any state $\rho$ of a $d$-level spin qudit in the form
\begin{align}
  \rho = \sum_{k=1}^{d^2} \rho_k^V Q_k^V,
  &&
  \rho_k^V \equiv \bk{Q^V_k}_\rho.
\end{align}
Given empirical estimates $E_j^V$ of the expectation values $\bk{\Pi_j^V}_\rho$, an empirical estimate $\tilde\rho$ of $\rho$ is then
\begin{align}
  \tilde\rho \equiv \sum_k \tilde\rho^V_k Q^V_k,
  \label{eq:reconstructed_state}
\end{align}
where
\begin{align}
  \tilde\rho^V_k \equiv \sum_j g^V_{kj} E_j^V
  \approx \sum_j g^V_{kj} \bk{\Pi_j^V}_\rho
  = \bk{Q^V_k}_\rho
  = \rho^V_k.
\end{align}
The eigenvectors of the overlap matrix $G_V$ provide strict upper bounds on the distance between the empirical estimate $\tilde\rho$ and the true state $\rho$.
If the estimates $E^V_j$ approximate $\bk{\Pi^V_j}_\rho$ to within an error $\epsilon$, i.e.
\begin{align}
  \abs{E^V_j-\bk{\Pi^V_j}_\rho} \le \epsilon,
\end{align}
then the error in the estimate $\tilde\rho^V_k$ of $\rho^V_k$ is
\begin{align}
  \abs{\tilde\rho^V_k-\rho^V_k}
  &= \abs{\sum_j g_{kj}^V \p{E^V_j-\bk{\Pi^V_j}_\rho}} \\
  &\le \sum_j \abs{g_{kj}^V \p{E^V_j-\bk{\Pi^V_j}_\rho}} \\
  &\le \epsilon \norm{\v g^V_k}_1,
\end{align}
where $\norm{\v g}_1=\sum_j\abs{g_j}$ is the 1-norm of the vector $\v g$.
Using the fact that the operators $Q^V_k$ are orthonormal, we can bound the 2-norm (Euclidean) distance $\norm{\tilde\rho-\rho}_2$ between $\tilde\rho$ and $\rho$ as
\begin{align}
  \norm{\tilde\rho-\rho}_2^2
  = \sum_k \abs{\tilde\rho^V_k - \rho^V_k}^2
  \le \epsilon^2 \N_V^2,
  \label{eq:bound}
\end{align}
where
\begin{align}
  \N_V^2 \equiv \min_U \sum_k \norm{U\v g^V_k}_1^2
  = \min_U \sum_k \f{\norm{U\v x^V_k}_1^2}{G^V_k}
  \label{eq:norm}
\end{align}
is a norm uniquely determined by the set of measurement axes $V$, with minimization performed over unitary transformations $U$ of all $\v g^V_k$ and $\v x^V_k$ to account for the fact that their 1-norms are generally basis-dependent.
Seen otherwise, this minimization accounts for the fact that we can perform the analysis between Eqs.~\eqref{eq:overlap_matrix} and \eqref{eq:bound} identically with any $G_V\to U G_V U^\dag$, where $U$ is an arbitrary unitary that mixes the projectors in $\v\Pi_V$.

@@

As a last point, we mention that the cost function $\tilde\N_V^2$ can also evaluate the quality of measurement axes chosen using existing methods for sparse signal recovery on the sphere \cite{rauhut2011sparse, alem2012sparse}.
In practice, we find that uniform random sampling outperforms these methods, in the sense that it generates low-cost measurement axes with higher probability.
Two observations may explain the poor performance of the sparse signal recovery methods: first, that spin qudit tomography essentially boils down to recovering sparse signals with a particular harmonic structure, namely with a fixed degree $\ell$.
The existing sparse signal recovery algorithms do not assume or account for any harmonic structure, and may therefore perform worse than expected on this particular subset of sparse signals.
Second, spin qudit tomography essentially requires reconstructing signals that are band-limited at degree $d$, but have $O(d)$ non-zero harmonic coefficients (out of $d^2$ coefficients in total).
The existing sparse signal recovery methods, on the other hand, may only work for signals with sparsity much smaller than $d$.
In any case, the fact that uniform sampling outperforms these methods only simplifies our randomized tomography protocol.

>>>>

For the practically minded, the mere existence of solutions to a problem is less interesting than the exposition of a particular solution.
On a high level, a spin qudit tomography protocol needs to
\begin{enumerate*}
\item select a set of measurement axes,
\item collect measurement data on spin projection onto these axes, and then
  \label{tomo:measure}
\item process the collected data to reconstruct the state of the spin qudit.
\end{enumerate*}
Whereas step \ref{tomo:measure} can involve a host of platform-dependent technical challenges, here we discuss the steps to take before and after collecting measurement data.

@@

A set of measurement axes $V=\set{\v v}$ nominally induces a set of projectors $\v\Pi_V=\set{\Pi_j^V}$ that will be measured in an experiment, where for shorthand we use a combined index $j\simeq\p{\v v_j,\mu_j}$ that specifies both a measurement axis $\v v_j$ and an outcome $\mu_j$, and for clarify we index objects by $V$ to emphasize their dependence on the choice of measurement axes.
The projectors $\v\Pi_V$ can be used to construct the overlap (Gram) matrix
\begin{align}
  G_V = \sum_{i,j=1}^{\abs{V}\times d} \obk{\Pi_i^V|\Pi_j^V} \op{i}{j}.
  \label{eq:overlap_matrix}
\end{align}
Eigenvectors of $G_V$ define linearly independent linear combinations of the projectors in $\v\Pi_V$.
If the projectors in $\v\Pi_V$ span the entire space of operators on a $d$-level spin qudit, which is a necessary and sufficient condition for $V$ to allow for full state tomography, then $G_V$ must have exactly $d^2$ nonzero eigenvalues.
Indexing these eigenvalues $G^V_k$ and the corresponding (normalized) eigenvectors $\v x^V_k\equiv(x^V_{k,1},x^V_{k,2},\cdots)$ by an integer $k\in\set{1,2,\cdots,d^2}$,
we can construct the orthonormal operators
\begin{align}
  Q^V_k \equiv \sum_j g^V_{kj} \Pi_j^V,
  &&
  g^V_{kj} \equiv \f{x^V_{kj}}{\sqrt{G^V_k}}.
\end{align}
These operators allow us to expand any state $\rho$ of a $d$-level spin qudit in the form
\begin{align}
  \rho = \sum_{k=1}^{d^2} \rho_k^V Q_k^V,
  &&
  \rho_k^V \equiv \bk{Q^V_k}_\rho.
\end{align}
Given empirical estimates $E_j^V$ of the expectation values $\bk{\Pi_j^V}_\rho$, an empirical estimate $\tilde\rho$ of $\rho$ is then
\begin{align}
  \tilde\rho \equiv \sum_k \tilde\rho^V_k Q^V_k,
  \label{eq:reconstructed_state}
\end{align}
where
\begin{align}
  \tilde\rho^V_k \equiv \sum_j g^V_{kj} E_j^V
  \approx \sum_j g^V_{kj} \bk{\Pi_j^V}_\rho
  = \bk{Q^V_k}_\rho
  = \rho^V_k.
\end{align}
The eigenvalues and eigenvectors of the overlap matrix $G_V$ provide strict upper bounds on the distance between the empirical estimate $\tilde\rho$ and the true state $\rho$.
If the estimates $E^V_j$ approximate $\bk{\Pi^V_j}_\rho$ to within an error $\epsilon$, i.e.
\begin{align}
  \abs{E^V_j-\bk{\Pi^V_j}_\rho} \le \epsilon,
\end{align}
then the error in the estimate $\tilde\rho^V_k$ of $\rho^V_k$ is
\begin{align}
  \abs{\tilde\rho^V_k-\rho^V_k}
  &= \abs{\sum_j g_{kj}^V \p{E^V_j-\bk{\Pi^V_j}_\rho}} \\
  &\le \sum_j \abs{g_{kj}^V \p{E^V_j-\bk{\Pi^V_j}_\rho}} \\
  &\le \epsilon \norm{\v g^V_k}_1,
\end{align}
where $\norm{\v g}_1=\sum_j\abs{g_j}$ is the 1-norm of the vector $\v g$.
Using the fact that the operators $Q^V_k$ are orthonormal, we can bound the 2-norm (Euclidean) distance $\norm{\tilde\rho-\rho}_2$ between $\tilde\rho$ and $\rho$ as
\begin{align}
  \norm{\tilde\rho-\rho}_2^2
  = \sum_k \abs{\tilde\rho^V_k - \rho^V_k}^2
  \le \epsilon^2 \N_V^2,
  \label{eq:bound}
\end{align}
where
\begin{align}
  \N_V^2 \equiv \min_U \sum_k \norm{U\v g^V_k}_1^2
  = \min_U \sum_k \f{\norm{U\v x^V_k}_1^2}{G^V_k}
  \label{eq:norm}
\end{align}
is a norm uniquely determined by the set of measurement axes $V$, with minimization performed over unitary transformations $U$ of all $\v g^V_k$ and $\v x^V_k$ to account for the fact that their 1-norms are generally basis-dependent.
Seen otherwise, this minimization accounts for the fact that we can perform the analysis between Eqs.~\eqref{eq:overlap_matrix} and \eqref{eq:bound} identically with any $G_V\to U G_V U^\dag$, where $U$ is an arbitrary unitary that mixes the projectors in $\v\Pi_V$.

@@

As a last point, we mention that the cost function $\tilde\N_V^2$ can also evaluate the quality of measurement axes chosen using existing methods for sparse signal recovery on the sphere \cite{rauhut2011sparse, alem2012sparse}.
In practice, we find that uniform random sampling outperforms these methods, in the sense that it generates low-cost measurement axes with higher probability.
Two observations may explain the poor performance of the sparse signal recovery methods: first, that spin qudit tomography essentially boils down to recovering sparse signals with a particular harmonic structure, namely with a fixed degree $\ell$.
The existing sparse signal recovery algorithms do not assume or account for any harmonic structure, and may therefore perform worse than expected on this particular subset of sparse signals.
Second, spin qudit tomography essentially requires recovering signals that are band-limited at degree $d$, but have $O(d)$ non-zero harmonic coefficients (out of $d^2$ coefficients in total).
The existing sparse signal recovery methods, on the other hand, may only perform well for signals with sparsity much smaller than $d$.
In any case, the fact that uniform sampling outperforms these methods only simplifies our randomized tomography protocol.
